\textbf{Proof Reconstruction in Classical Propositional Logic}

### 2017-06-16 slides/Jonathan-Proof-Reconstruction.tex
[ slides ] Renamed slides. Fixed atp-conjunct explanation.

% --------------------------------------------------------------------------
% Tikz Configuration
% --------------------------------------------------------------------------
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usepackage{rotating}

@@

@@

\vfill
\begin{itemize}
\item Definition
\begin{equation*}
\text{\color{plum!40!black}conjunct}(ϕ₁ ∧ ⋯ ∧ \phi_n, \omega)
= \begin{cases}
\phi_{i} \ \ \ \ \text{if }\omega\text{ is equal to some }\phi_i\\
\phi    \ \ \ \ \text{otherwise}
\end{cases}
\end{equation*}

\pause
\item Inference rules involved
\begin{center}
\begin{columns}
\begin{column}{0.45\textwidth}

 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{\scriptsize $∧$-\texttt{proj}$₁$}
  \UnaryInfC{$ϕ$}
  \end{prooftree}
\end{column}

\begin{column}{0.45\textwidth}
 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{\scriptsize $∧$-\texttt{proj}$₂$}
  \UnaryInfC{$ψ$}
  \end{prooftree}
\end{column}

\begin{column}{0.1\textwidth}  %%<--- here
   \begin{tikzpicture}[overlay, remember picture, scale=2]
% \tikzset{shift={(current page.north)}, xshift=cm,yshift=-2cm}
\node[vertex, fill=black] (d1) {};
\node[vertex, fill=black, below left=0.2cm and 0.1cm of d1] (d2) {};
\node[vertex, fill=black, below right=0.2cm and 0.1cm of d1] (d3) {};

\node[vertex, fill=black, below left=0.2cm and 0.1cm of d3] (q) {};
\node[vertex, fill=black, below right=0.2cm and 0.1cm of d3] (w) {};

\node[vertex, fill=black, below right=0.2cm and 0.1cm of q] (s) {};

\draw[-, thick] (d1) to (d2);
\draw[-, thick] (d1) to (d3);
\draw[-, thick] (d3) to (q);
\draw[-, thick] (d3) to (w);
\draw[-, thick] (q) to (s);
\end{tikzpicture}
\end{column}
\end{columns}
\end{center}
\vskip 2mm

\item Target example
\[ φ := (¬ p ∧ (¬ q ⇔ ¬ r)) ∧ (p ∧ (¬ q ⇔ ¬r))\]

@@

\item
\begin{minted}{cagda}
 conjunct(φ , ¬ q) ≡ ϕ
\end{minted}
\vskip 3mm
\item
\begin{minted}{cagda}
 conjunct(φ , ¬ q ⇔ ¬ r) ≡ ¬ q ⇔ ¬ r
\end{minted}
\vskip 3mm

\item
\begin{minted}{cagda}
 conjunct(φ , p)   ≡ p
\end{minted}

\end{itemize}

@@

\begin{frame}[fragile, label=atp-conjunct]

@@

conjunct-path : (φ : Prop) → (ω : Prop) → Path → Path
conjunct-path φ ω path with ⌊ eq φ ω ⌋
... | true  = path ∷ʳ pick
... | false with conj-view φ
...   | other _   = []
...   | conj φ₁ φ₂ with conjunct-path φ₁ ω []
...     | subpath@(_ ∷ _) = (path ∷ʳ proj₁) ++ subpath
...     | [] with conjunct-path φ₂ ω []
...         | subpath@(_ ∷ _) = (path ∷ʳ proj₂) ++ subpath
...         | []              = []
\end{minted}

@@

\begin{frame}[fragile, label=atp-conjunct-proof]{The \texttt{conjunct} function and its theorem, \texttt{atp-conjunct}}
\vfill
\begin{minted}{cagda}
conjunct : (φ : Prop) → (ω : Prop) → Prop
conjunct φ ω with conj-view φ | conjunct-path φ ω []
... | _          | []           = φ
... | conj φ₁ φ₂ | pick  ∷ path = φ
... | conj φ₁ φ₂ | proj₁ ∷ path = conjunct φ₁ ω
... | conj φ₁ φ₂ | proj₂ ∷ path = conjunct φ₂ ω
... | other .φ   | _            = φ

atp-conjunct
  : ∀ {Γ} {φ}
  → (ω : Prop)
  → Γ ⊢ φ
  → Γ ⊢ conjunct φ ω

atp-conjunct {Γ} {φ} ω Γ⊢φ
  with conj-view φ | conjunct-path φ ω []
...  | _          | []           = Γ⊢φ
...  | conj φ₁ φ₂ | pick  ∷ path = Γ⊢φ
...  | conj φ₁ φ₂ | proj₁ ∷ path = atp-conjunct ω (∧-proj₁ Γ⊢φ)
...  | conj φ₁ φ₂ | proj₂ ∷ path = atp-conjunct ω (∧-proj₂ Γ⊢φ)
...  | other .φ   | (_ ∷ _)      = Γ⊢φ
\end{minted}
\vfill
\gobackreconstruction
\end{frame}

@@

\begin{itemize}
  \item The problem is $p ∧ q ⊢ q ∧ p$
  \item In \texttt{TPTP} format
\begin{tptp}
$ cat problem.tptp
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
\end{tptp}
  \item How to use Athena with your problem
  \vspace*{5mm}
\begin{tikzpicture}
\node[rectangle, text=white,fill=black, align=left, font=\small]
{
\texttt{\$ metis --show proof problem.tptp > problem.tstp}\hspace{1cm}\\
\texttt{\$ athena problem.tstp}\\
\texttt{\$ agda problem.agda}
  };
\end{tikzpicture}

\begin{frame}[fragile, allowframebreaks]{\texttt{TSTP} Derivation}
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
fof(subgoal_0, plain, q,
    inference(strip, [], [goal])).
fof(subgoal_1, plain, q => p,
    inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ q,
inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, (~ q),
    inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p & q,
    inference(canonicalize, [], [a])).
fof(normalize_0_2, plain, q,
    inference(conjunct, [], [normalize_0_1])).
fof(normalize_0_3, plain, $false,
    inference(simplify, [],
        [normalize_0_0, normalize_0_2])).
cnf(refute_0_0, plain, $false,
    inference(canonicalize, [], [normalize_0_3])).
fof(negate_1_0, plain, ~ (q => p),
    inference(negate, [], [subgoal_1])).
fof(normalize_1_0, plain, ~ p & q,
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, p & q,
     inference(canonicalize, [], [a])).
fof(normalize_1_2, plain, p,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, q,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_4, plain, $false,
    inference(simplify, [],
      [normalize_1_0, normalize_1_2, normalize_1_3])).
cnf(refute_1_0, plain, ($false),
    inference(canonicalize, [], [normalize_1_4])).
\end{tptp}
\end{frame}

\begin{frame}[fragile, label=verified-example]{$p ∧ q ⊢ q ∧ p$}{Reconstructed Proof}
\vfill
\begin{minted}{cagda}
p, q, a, goal, subgoal₀, subgoal₁ : Prop

@@

\begin{frame}[fragile, label=verified-example-2]{$p ∧ q ⊢ q ∧ p$}{Reconstructed Proof}
\vfill
\begin{minted}{cagda}
a : Prop
a = (p ∧ q)

@@

\begin{frame}[fragile, label=verified-example-3]{$p ∧ q ⊢ q ∧ p$}{Reconstructed Proof}
\vfill
\begin{minted}{cagda}
subgoal₁ : Prop
subgoal₁ = (q ⇒ p)

@@

\begin{frame}[fragile, label=verified-example-4]{$p ∧ q ⊢ q ∧ p$}{Reconstructed proof}
\vfill
\begin{minted}{cagda}
-- Premise.
Γ = [ a ]

@@

\begin{frame}[label=contributions]{Contributions}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll|}
\hline
\textbf{Name} & \textbf{References} \\
\hline
\href{http://github.com/jonaprieto/agda-metis}{\texttt{Agda-Metis}}   & \citep{AgdaMetis}\\
\href{http://github.com/jonaprieto/agda-prop}{\texttt{Agda-Prop}}     & \citep{AgdaProp}\\
\href{http://github.com/jonaprieto/athena}{\texttt{Athena}}           & \citep{Athena}\\
\href{http://github.com/jonaprieto/online-atps}{\texttt{OnlineATPs}}  & \citep{OnlineATPs}\\
\href{http://github.com/jonaprieto/prop-pack}{\texttt{Prop-Pack}}     & \citep{ProPack}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}
\vfill
\end{frame}

>>>>

% --------------------------------------------------------------------------
% Tikz Configuration
% --------------------------------------------------------------------------
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{automata}
\usetikzlibrary{calc}
\usepackage{rotating}

@@

\begin{frame}[fragile]{Proof Reconstruction: Overview}
\vfill
\begin{tikzpicture}
\node(problem) at (0,0)
   {\includegraphics[scale=0.7]{figures/problem}};
\node[right = 0.8cm of problem, hyperlink node=tptp-syntax](tptp){
\includegraphics[scale=0.7]{figures/tptp}
};
\node[right= 0.8cm of tptp, hyperlink node=metis] (metis)
  {\includegraphics[scale=0.7]{figures/metis}};
\node[right= 0.8cm of metis, hyperlink node=tstp-syntax] (tstp)
  {\includegraphics[scale=0.7]{figures/tstp}};
\node[below= 1.5cm of tstp, hyperlink node=athena] (athena)
  {\includegraphics[scale=0.7]{figures/athena}};
\node[left = 0.6cm of athena, hyperlink node=agda-file] (agdafile)
  {\includegraphics[scale=0.7]{figures/agdafile}};
\node[left = 0.5cm of agdafile] (agda)
  {\includegraphics[scale=0.7]{figures/agda}};
\node[below = 0.18cm of problem, hyperlink node=verified-example] (verified)
  {\includegraphics[scale=0.7]{figures/verified}};
\node[below = 0.18cm of verified, opacity=0.5, hyperlink node=failure-example] (failure)
  {\includegraphics[scale=0.7]{figures/failure}};

@@

\vfill
\begin{itemize}
\item Definition
\begin{equation*}
\text{\color{plum!40!black}conjunct}(\overbrace{ϕ₁ ∧ ⋯ ∧ \phi_n}^{\phi}, \psi)
= \begin{cases}
\phi_{i} \ \ \ \ \text{if }\psi\text{ is equal to some }\phi_i\\
\phi    \ \ \ \ \text{otherwise}
\end{cases}
\end{equation*}

\pause
\item Inference rules involved
\begin{center}
\begin{columns}
\begin{column}{0.42\textwidth}

 \begin{prooftree}
  \AxiomC{$ϕ₁ ∧ φ₂$}
  \RightLabel{\scriptsize $∧$-\texttt{proj}$₁$}
  \UnaryInfC{$ϕ₁$}
  \end{prooftree}
\end{column}

\begin{column}{0.42\textwidth}
 \begin{prooftree}
  \AxiomC{$ϕ₁ ∧ φ₂$}
  \RightLabel{\scriptsize $∧$-\texttt{proj}$₂$}
  \UnaryInfC{$φ₂$}
  \end{prooftree}
\end{column}

\begin{column}{0.2\textwidth}  %%<--- here
  \begin{tikzpicture}[
    ->
  , >=stealth'
  , shorten >=1pt
  , auto
%  , node distance=2.8cm
  , thick
  , every state/.style={fill=plum!45!black,draw=none,text=white}
  , overlay
  , remember picture
    , transform canvas={scale=.75}
    ]
  \tikzset{shift={(current page.center)}, xshift=4.5cm,yshift=1cm}

  \node[state]  (phi)                     {$φ$};
  \node[state]  (phi1) [below left =1.3cm and 0.1cm of phi]  {$φ₁$};
  \node[state]  (phi2) [below right=1.3cm and 0.1cm of phi]  {$φ₂$};
  \node[state]  (phi3)  [below left =1.3cm and 0.1cm of phi2]  {$φ₃$};
  \node[state]  (phi4)  [below right =1.3cm and 0.1cm of phi2]  {$φ₄$};

  \path[every node/.style={sloped,anchor=south,auto=false}]
        (phi) edge node {\tiny $∧$-\texttt{proj}$₁$} (phi1)            
        (phi) edge node {\tiny $∧$-\texttt{proj}$₂$} (phi2)
        (phi2) edge node {\tiny $∧$-\texttt{proj}$₁$} (phi3)
        (phi2) edge node {\tiny $∧$-\texttt{proj}$₂$} (phi4);

\end{tikzpicture}
\end{column}
\end{columns}
\end{center}
\vskip 2mm

\item Example
\[ φ := φ₁ ∧ \overbrace{(φ₃ ∧ φ₄)}^{φ₂}\]

@@

\item
\texttt{conjunct}$(φ , φ₃ ∧ φ₁) ≡ ϕ$
\vskip 3mm
\item
\texttt{conjunct}$(φ , φ₃)      ≡ ϕ₃$
\vskip 3mm
\item
\texttt{conjunct}$(φ , φ₂)      ≡ ϕ₂$
\vskip 3mm
\end{itemize}

@@

\begin{frame}[fragile]

@@

conjunct-path : (φ ψ : Prop) → Path → Path
conjunct-path φ ψ path with ⌊ eq φ ψ ⌋
... | true  = path ∷ʳ pick
... | false with conj-view φ
...   | other _   = []
...   | conj φ₁ φ₂ with conjunct-path φ₁ ψ []
...     | subpath@(_ ∷ _) = (path ∷ʳ proj₁) ++ subpath
...     | [] with conjunct-path φ₂ ψ []
...         | subpath@(_ ∷ _) = (path ∷ʳ proj₂) ++ subpath
...         | []              = []
\end{minted}

@@

\begin{frame}[fragile, label=atp-conjunct-proof]{The \texttt{conjunct} function and its theorem, \texttt{atp-conjunct}}
\vfill
\begin{minted}{cagda}
conjunct : (φ ψ : Prop) → Prop
conjunct φ ψ with conj-view φ | conjunct-path φ ψ []
... | _          | []           = φ
... | conj _ _ | pick  ∷ path = φ
... | conj φ₁ _ | proj₁ ∷ path = conjunct φ₁ ψ
... | conj _ φ₂ | proj₂ ∷ path = conjunct φ₂ ψ
... | other .φ   | _            = φ

atp-conjunct
  : ∀ {Γ} {φ}
  → Γ ⊢ φ
  → (ψ : Prop)
  → Γ ⊢ conjunct φ ψ

atp-conjunct {Γ} {φ} Γ⊢φ ψ
  with conj-view φ | conjunct-path φ ψ []
...  | _          | []           = Γ⊢φ
...  | conj φ₁ φ₂ | pick  ∷ path = Γ⊢φ
...  | conj φ₁ φ₂ | proj₁ ∷ path = atp-conjunct ψ (∧-proj₁ Γ⊢φ)
...  | conj φ₁ φ₂ | proj₂ ∷ path = atp-conjunct ψ (∧-proj₂ Γ⊢φ)
...  | other .φ   | (_ ∷ _)      = Γ⊢φ
\end{minted}
\vfill
\gobackreconstruction
\end{frame}

@@

\begin{itemize}
  \item The problem is $p ∧ q ⊢ q ∧ p$
  \item In \texttt{TPTP} format
\begin{tptp}
$ cat problem.tptp
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
\end{tptp}
  \item How to use Athena with your problem

\begin{tptp}
$ metis --show proof problem.tptp > problem.tstp
$ athena problem.tstp
$ agda problem.tstp
\end{tptp}
\end{itemize}
\end{frame}

\begin{frame}[fragile, allowframebreaks]{\texttt{TSTP} Derivation given by \texttt{Metis}}
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
fof(subgoal_0, plain, q,
    inference(strip, [], [goal])).
fof(subgoal_1, plain, q => p,
    inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ q,
inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, (~ q),
    inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p & q,
    inference(canonicalize, [], [a])).
fof(normalize_0_2, plain, q,
    inference(conjunct, [], [normalize_0_1])).
fof(normalize_0_3, plain, $false,
    inference(simplify, [],
        [normalize_0_0, normalize_0_2])).
cnf(refute_0_0, plain, $false,
    inference(canonicalize, [], [normalize_0_3])).
fof(negate_1_0, plain, ~ (q => p),
    inference(negate, [], [subgoal_1])).
fof(normalize_1_0, plain, ~ p & q,
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, p & q,
     inference(canonicalize, [], [a])).
fof(normalize_1_2, plain, p,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, q,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_4, plain, $false,
    inference(simplify, [],
      [normalize_1_0, normalize_1_2, normalize_1_3])).
cnf(refute_1_0, plain, ($false),
    inference(canonicalize, [], [normalize_1_4])).
\end{tptp}
\end{frame}

\begin{frame}[fragile, label=verified-example]{Problem $p ∧ q ⊢ q ∧ p$}{Reconstructed Proof generated with \texttt{Athena}}
\vfill
\begin{minted}{cagda}
p, q, a, goal, subgoal₀, subgoal₁ : Prop

@@

\begin{frame}[fragile, label=verified-example-2]{Problem $p ∧ q ⊢ q ∧ p$}{Reconstructed Proof}
\vfill
\begin{minted}{cagda}
a : Prop
a = (p ∧ q)

@@

\begin{frame}[fragile, label=verified-example-3]{Problem $p ∧ q ⊢ q ∧ p$}{Reconstructed Proof}
\vfill
\begin{minted}{cagda}
subgoal₁ : Prop
subgoal₁ = (q ⇒ p)

@@

\begin{frame}[fragile, label=verified-example-4]{Problem $p ∧ q ⊢ q ∧ p$}{Reconstructed proof}
\vfill
\begin{minted}{cagda}
-- Premise.
Γ = [ a ]

@@

\begin{frame}[label=contributions]{Contributions}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll|}
\hline
\textbf{Name} & \textbf{References} \\
\hline
\href{http://github.com/jonaprieto/agda-metis}{\texttt{Agda-Metis}}   & \citep{AgdaMetis}\\
\href{http://github.com/jonaprieto/agda-prop}{\texttt{Agda-Prop}}     & \citep{AgdaProp}\\
\href{http://github.com/jonaprieto/athena}{\texttt{Athena}}           & \citep{Athena}\\
\href{http://github.com/jonaprieto/online-atps}{\texttt{OnlineATPs}}  & \citep{OnlineATPs}\\
\href{http://github.com/jonaprieto/prop-pack}{\texttt{Prop-Pack}}     & \citep{ProPack}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\end{table}
\vfill
\end{frame}
